sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2022-11-02 doc/paper-jss/paper-jss.Rmd
Edits

---
documentclass: jss
author:
  - name: Sean C. Anderson
    orcid: 0000-0001-9563-1937
    affiliation: 'Pacific Biological Station, Fisheries and Oceans Canada \AND'
    address: |
      | Pacific Biological Station, Nanaimo, BC, Canada
    email: \email{sean.anderson@dfo-mpo.gc.ca}
  - name: Eric J. Ward
    orcid: 0000-0002-4359-0296
    affiliation: 'Northwest Fisheries Science Center, National Marine Fisheries Service \AND'
  - name: Philina A. English
    orcid: 0000-0003-2992-6782
    address: ''
    affiliation: 'Pacific Biological Station, Fisheries and Oceans Canada \AND'
  - name: Lewis A. K. Barnett
    orcid: 0000-0002-9381-8375
    affiliation: 'Alaska Fisheries Science Center, National Marine Fisheries Service'
bibliography: refs.bib
title:
  formatted: "\\pkg{sdmTMB}: An \\proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # If you use tex in the formatted title, also supply version without
  plain:     "sdmTMB: An R Package for Fast, Flexible, and Accessible Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # For running headers, if needed
  short:     "\\pkg{sdmTMB}: geostatistical SPDE-based GLMMs with \\pkg{TMB}"
abstract: >
  Geostatistical data are common across disciplines and available in increasingly large volumes. However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to implement such that spatial autocorrelation is frequently ignored, or addressed in less interpretable ways. Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of `stats::glm()`, \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach. SPDE matrices are constructed with \pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}), although penalised likelihood and Bayesian inference (via \pkg{tmbstan} and \pkg{rstan}) are also possible. We describe the statistical model, illustrate the package through two case studies, and compare the functionality, speed, and interface to related software. The case studies illustrate \pkg{sdmTMB}'s flexibility and useability by implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (asymmetry in spatial correlations). Modelling of these common but computationally complex relationships is further aided by built-in handling of delta (hurdle) models, break-point effects, and diagnostic tools including out-of-sample cross validation. Finally, we demonstrate that \pkg{sdmTMB} is 4--16 times faster than \pkg{INLA} or \pkg{inlabru}. \pkg{sdmTMB}'s accessible interface has already opened this useful class of models to a wider audience within ecological species distribution modelling. We anticipate its application to the full spectrum geostatistical data will prove equally fruitful.
keywords:
  # at least one keyword must be supplied
  formatted: [test]
  plain:     [test]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{bm}
  \usepackage{lscape}
  \usepackage{tablefootnote}
  \usepackage{threeparttable}
  \usepackage{booktabs}
output: rticles::jss_article
---

@@

Ecological data are often collected in space or in space repeatedly over time.
While such data are a rich source of information about ecological processes [@legendre1989a; @rossi1992; @tilman1997spatial], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart due to measured and unmeasured variables [@cressie1993; @diggle2007; @cressie2011].
While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
Accounting for this residual correlation is important because doing so allows for valid statistical inference [@legendre1989a; @dormann2007], can improve predictions  [e.g., @shelton2014], and can be of ecological interest itself by, for example, identifying locations with similar population responses [e.g., @thorson2019d; @barnett2021].
<!-- LB: seems like we needed some explicit example of why it would be of ecological interest itself to those who are unfamiliar, so I added allusion to local trends...but maybe also mentioning quantification of patchiness or something similar?. More generally, it might be good to expand on the utility of such models for applied ecology, if we can find room within the tight word limit? -->

@@

Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which implements geostatistical spatial and spatiotemporal GLMMs using \pkg{TMB} for model fitting and \pkg{INLA} to set up SPDE matrices.
Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} [@wood2017a], for a specific class of spatial and spatiotemporal models.
One common application in the field of ecology is species distribution models (SDMs), hence the package name.
This paper describes the basic functionality of this \proglang{R} package and its underlying statistical model, illustrates its use through two case studies, and concludes with a comparison to related software.

>>>>

---
documentclass: jss
author:
  - name: Sean C. Anderson
    orcid: 0000-0001-9563-1937
    affiliation: 'Pacific Biological Station, Fisheries and Oceans Canada \AND'
    address: |
      | Pacific Biological Station, Nanaimo, BC, Canada
    email: \email{sean.anderson@dfo-mpo.gc.ca}
  - name: Eric J. Ward
    orcid: 0000-0002-4359-0296
    affiliation: 'Northwest Fisheries Science Center, National Marine Fisheries Service \AND'
  - name: Philina A. English
    orcid: 0000-0003-2992-6782
    address: ''
    affiliation: 'Pacific Biological Station, Fisheries and Oceans Canada \AND'
  - name: Lewis A. K. Barnett
    orcid: 0000-0002-9381-8375
    affiliation: 'Alaska Fisheries Science Center, National Marine Fisheries Service'
bibliography: refs.bib
title:
  formatted: "\\pkg{sdmTMB}: An \\proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # If you use tex in the formatted title, also supply version without
  plain:     "sdmTMB: An R Package for Fast, Flexible, and Accessible Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # For running headers, if needed
  short:     "\\pkg{sdmTMB}: geostatistical SPDE-based GLMMs with \\pkg{TMB}"
abstract: >
  Geostatistical data are pervasive and available in increasingly large volumes. However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to implement. Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach. SPDE matrices are constructed with \pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}), although penalised likelihood and Bayesian inference (via \pkg{tmbstan} and \pkg{rstan}) are possible. We describe the statistical model, illustrate the package through two case studies, and compare the functionality, speed, and interface to related software. The case studies illustrate \pkg{sdmTMB}'s flexibility and useability by implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (asymmetry in spatial correlations). Additional functionality includes delta (hurdle) models, break-point effects, and out-of-sample cross validation. Finally, we demonstrate that \pkg{sdmTMB} is 4--16 times faster than \pkg{INLA} or \pkg{inlabru}. \pkg{sdmTMB}'s accessible interface has already opened this useful class of models to a wider audience within ecological species distribution modelling (SDMs). We anticipate its application to the full spectrum of geostatistical data will prove equally fruitful.
keywords:
  # at least one keyword must be supplied
  formatted: [test]
  plain:     [test]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{bm}
  \usepackage{lscape}
  \usepackage{tablefootnote}
  \usepackage{threeparttable}
  \usepackage{booktabs}
output: rticles::jss_article
---

@@

Data are often collected in space or in space repeatedly over time.
While such data are a rich source of information about mechanistic processes [@legendre1989a; @rossi1992; @tilman1997spatial], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart due to measured and unmeasured variables [@cressie1993; @diggle2007; @cressie2011].
While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
Accounting for this residual correlation is important because doing so allows for valid statistical inference [@legendre1989a; @dormann2007], can improve predictions  [e.g., @shelton2014], and can be of interest itself by, for example, identifying locations with shared responses [e.g., @thorson2019d; @barnett2021].
<!-- LB: seems like we needed some explicit example of why it would be of ecological interest itself to those who are unfamiliar, so I added allusion to local trends...but maybe also mentioning quantification of patchiness or something similar?. More generally, it might be good to expand on the utility of such models for applied ecology, if we can find room within the tight word limit? -->

@@

Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which implements geostatistical spatial and spatiotemporal GLMMs using \pkg{TMB} for model fitting and \pkg{INLA} to set up SPDE matrices.
Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} [@wood2017a], for a specific class of spatial and spatiotemporal models.
One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \pkg{TMB}: sdmTMB).
This paper describes the basic functionality of this \proglang{R} package and its underlying statistical model, illustrates its use through two case studies, and concludes with a comparison to related software.

### 2022-11-02 doc/paper-jss/pcod.Rmd
Edits

@@

```{r packages, message=FALSE, warning=FALSE, cache=FALSE}
library(ggplot2)
library(dplyr)
library(sdmTMB)
theme_set(theme_minimal())
```

We will fit a geostatistical spatiotemporal model with sdmTMB for the purposes of index standardization. We will use a data set built into the package: trawl survey data for Pacific Cod in Queen Charlotte Sound. The density units are kg/km^2^. Here, X and Y are UTM zone 9 coordinates in kilometres. If our spatial coordinates were not already in an equidistant projection, we would transform them [e.g., using `sf::st_transform()`, @pebesma2018] to ensure that distance remains constant throughout the study region or using the built-in `sdmTMB::add_utm_columns()` function.

@@

Next, we want to predict on a fine-scale grid over the entire survey domain so we can sum up the expected biomass across the entire region each year. There is a grid built into the sdmTMB package for Queen Charlotte Sound (named `qcs_grid`). Our prediction grid also needs to have all the covariates that we used in the model above. In this case, that is a `year` column as well as `X` and `Y`.

>>>>

\clearpage

@@

```{r packages, message=FALSE, warning=FALSE, cache=FALSE}
library("ggplot2")
library("dplyr")
library("sdmTMB")
theme_set(theme_minimal())
```

We will fit a geostatistical spatiotemporal model with \pkg{sdmTMB} for the purposes of index standardization. We will use a data set built into the package: trawl survey data for Pacific Cod in Queen Charlotte Sound. The density units are kg/km^2^. Here, X and Y are UTM zone 9 coordinates in kilometres. If our spatial coordinates were not already in an equidistant projection, we would transform them [e.g., using `sf::st_transform()`, @pebesma2018] to ensure that distance remains constant throughout the study region or using the built-in `sdmTMB::add_utm_columns()` function.

@@

Next, we want to predict on a fine-scale grid over the entire survey domain so we can sum up the expected biomass across the entire region each year. There is a grid built into the \pkg{sdmTMB} package for Queen Charlotte Sound (named `qcs_grid`). Our prediction grid also needs to have all the covariates that we used in the model above. In this case, that is a `year` column as well as `X` and `Y`.
